---
title: "EDA: Maximzing WOBA with Launch Angle and Exit Velocity"
author: "Sarah Sult, Brooke Coneeny, Erin Franke"
date: '`r Sys.Date()`'
output:
  xaringan::moon_reader:
    css: xaringan-themer.css
    nature:
      slideNumberFormat: "%current%"
      highlightStyle: github
      highlightLines: true
      ratio: 16:9
      countIncrementalSlides: true
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(fig.retina = 3, warning = FALSE, message = FALSE)
```

```{r xaringan-themer, include=FALSE, warning=FALSE}
library(xaringanthemer)
style_mono_accent(base_color = "#cc002b")
```

# Motivation

- Two of the most popular stats in the MLB have recently been launch angle and exit velocity 
  - We wondered: can we prescribe a woba maximizing launch angle for an individual player 
  based on their power profile?
```{r, include=FALSE}
library(baseballr)
library(tidyverse)
library(mgcv)
batter_all_2020 <- read_rds("../private_data/all2020data.rds")
batter_all_2019 <- read_rds("../private_data/all2019data.rds")
batter_all_2021 <- read_rds("../private_data/all2021data.rds")
batter_all <- rbind(batter_all_2019, batter_all_2020, batter_all_2021)

#get all balls in play  
in_play <- batter_all %>%
  filter(description == "hit_into_play") %>%
  select(launch_angle, woba_value, player_name, launch_speed) 

#max EV per player and combine with balls in play (so there is a column categorizing the player who 
  #hit the ball by their max EV in the season)
max_EV_grouped2 <- batter_all %>%
  group_by(player_name) %>%
  summarize(max_EV = max(launch_speed, na.rm = TRUE)) %>%
  filter(max_EV < 200 & max_EV>0) %>%
  right_join(in_play, by = c("player_name")) %>%
  mutate(EV_group = cut(max_EV, breaks = c(0, 110, 150), 
                        labels = c("<110 EV Hitters", "110+ EV Hitters")))

#group by 3 EVs instead
max_EV_grouped3 <- batter_all %>%
  group_by(player_name) %>%
  summarize(max_EV = max(launch_speed, na.rm = TRUE)) %>%
  filter(max_EV < 200 & max_EV>0) %>%
  right_join(in_play, by = c("player_name")) %>%
  mutate(EV_group = cut(max_EV, breaks = c(0, 105.99, 111.99, 150), 
                        labels = c("<106 EV Hitters", "106+ EV Hitters", "112+ EV Hitters")))

#graphed divided into 2 EV groups
wOBA_angle_velo_graph_2 <- max_EV_grouped2 %>%
  ggplot(aes(x=launch_angle, y = woba_value, color = EV_group)) +
  geom_smooth(method = "loess", span = .2, se = FALSE)+
  coord_cartesian(ylim=c(0,1), xlim=c(-40,60)) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0))+
  scale_color_manual(values = c("red", "blue"))+
  theme_bw()+
  theme(legend.title = element_blank())+
  labs(x= "Launch Angle",
       y = "wOBA",
       title = "wOBA by Launch Angle")

#grouped into 3 EV 
wOBA_angle_velo_graph_3 <- max_EV_grouped3 %>%
  ggplot(aes(x=launch_angle, y = woba_value, color = EV_group)) +
  geom_smooth(method = "loess", span = .2, se = FALSE)+
  coord_cartesian(ylim=c(0,1), xlim=c(-40,60)) +
  scale_x_continuous(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0))+
  scale_color_manual(values = c("black", "orange", "blue"))+
  theme_bw()+
  theme(legend.title = element_blank())+
  labs(x= "Launch Angle",
       y = "wOBA",
       title = "wOBA by Launch Angle")
```

.pull-left[

```{r, echo=FALSE, fig.height=4}
wOBA_angle_velo_graph_2 
```

]
.pull-right[

```{r, echo = FALSE, fig.height=4}
wOBA_angle_velo_graph_3
```
]

---
# Data

- Most of our EDA graphs use 2019 MLB data, the last 162 game season
- Data is taken from Statcast on Baseball Savant
- Variables
  - We are interested in modeling **woba** using **launch angle** and **exit velocity**
  - Other relevant variables we looked at during the EDA process include **battter handedness, 
  pitcher handedness, infield alignment, pitch type, at-bat outcome, location of where the ball was fielded**
  
```{r out.width='30%', echo = FALSE, fig.align='center'}
data_example <- batter_all_2019 %>%
  filter(description == "hit_into_play") %>%
  rename(`player` = player_name, `exit velocity` = launch_speed, `launch angle` = launch_angle, wOBA = woba_value, outcome = events) %>%
  select(player, `exit velocity`, `launch angle`, wOBA, outcome)
knitr::kable(head(data_example), format = 'html')
```
---
# Hit outcome by launch angle and exit velocity 

```{r, fig.width = 14, fig.height = 7, echo = FALSE, fig.align='center'}
library(nationalparkcolors)
pal2 <- park_palette("Everglades")

batter_all_2019 %>%
  filter(description == "hit_into_play") %>%
  mutate(events_group = case_when(
    events %in% c("field_out", "other_out", "grounded_into_double_play", "double_play", 
                  "fielders_choice_out", "force_out", "sac_fly", "sac_fly_double_play", "sac_bunt_double_play", 
                  "field_error", "sac_fly", "fielders_choice", "triple_play") ~ "out",
    events == "single" ~ "single",
    events == "double" ~ "double", 
    events == "triple" ~ "triple", 
    events == "home_run" ~ "home run", 
    TRUE ~ "other")) %>%
  filter(events_group %in% c("out", "single", "double", "triple", "home run")) %>%
  ggplot(aes(x=launch_angle, y = launch_speed, color = events_group)) +
  geom_point(alpha = 0.75) + 
  scale_x_continuous(n.breaks = 15, limits = c(-60, 90))+
  scale_color_manual(values = rev(pal2))+
  labs(x = "launch angle", y = "exit velocity (mph)", color = "")+
  theme_minimal() +
  theme(panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank())
```
---
#Breaking down the previous graph proportionally...

.pull-left[
```{r, fig.width = 7, fig.height = 5, echo = FALSE, fig.align='center'}
launch_angle_outcomes <- batter_all_2019 %>%
  filter(description == "hit_into_play") %>%
  mutate(
    events_group = case_when(
      events %in% c("field_out", "other_out", "grounded_into_double_play", "double_play", 
                    "fielders_choice_out", "force_out", "sac_fly", "sac_fly_double_play", "sac_bunt_double_play", 
                    "field_error", "sac_fly", "fielders_choice", "triple_play") ~ "out",
      events == "single" ~ "single",
      events == "double" ~ "double", 
      events == "triple" ~ "triple", 
      events == "home_run" ~ "home run", 
      TRUE ~ "other"),
    launch_angle_group = cut(
      launch_angle, breaks = c(-20, -15, -10, -5, 0, 5, 10, 15, 20, 25, 
                                        30, 35, 40, 45, 50))) %>%
  filter(
    events_group %in% c("out", "single", "double", "triple", "home run"),
    !is.na(launch_angle_group)
  ) %>%
  select(events_group, launch_angle_group) %>%
  ggplot(mapping = aes(x = launch_angle_group, fill = events_group)) +
  geom_bar(position = "dodge") +
  scale_fill_manual(values = pal2)+
  theme_classic() +
  theme(
    axis.title.x = element_blank(),
    legend.position = "none"
  )
launch_angle_outcomes
```
]

.pull-right[
```{r, fig.width = 7, fig.height = 5, echo = FALSE, fig.align='center'}
batter_all_2019 %>%
  filter(description == "hit_into_play") %>%
  mutate(events_group = case_when(
    events %in% c("field_out", "other_out", "grounded_into_double_play", "double_play", 
                  "fielders_choice_out", "force_out", "sac_fly", "sac_fly_double_play", "sac_bunt_double_play", 
                  "field_error", "sac_fly", "fielders_choice", "triple_play") ~ "out",
    events == "single" ~ "single",
    events == "double" ~ "double", 
    events == "triple" ~ "triple", 
    events == "home_run" ~ "home run", 
    TRUE ~ "other")) %>%
  filter(events_group %in% c("out", "single", "double", "triple", "home run"), 
         launch_angle >= -20 & launch_angle <=50) %>%
  mutate(launch_angle_group = 
           cut(launch_angle, breaks = c(-20, -15, -10, -5, 0, 5, 10, 15, 20, 25, 
                                        30, 35, 40, 45, 50))) %>%
  filter(!is.na(launch_angle_group)) %>%
  ggplot(aes(x = launch_angle_group, fill = events_group))+
  geom_bar(position = "fill")+
  scale_fill_manual(values = pal2)+
  labs(x = "launch angle", y= "", fill = "")+
  theme_minimal()
```
]
---
## Infield alignment did not impact the distribution of events at each angle grouping

```{r, fig.width = 15, fig.height = 6, echo = FALSE, fig.align='center'}
faceted_by_alignment <- batter_all_2019 %>%
  filter(description == "hit_into_play") %>%
  mutate(events_group = case_when(
    events %in% c("field_out", "other_out", "grounded_into_double_play", "double_play", 
                  "fielders_choice_out", "force_out", "sac_fly", "sac_fly_double_play", "sac_bunt_double_play", 
                  "field_error", "sac_fly", "fielders_choice", "triple_play") ~ "out",
    events == "single" ~ "single",
    events == "double" ~ "double", 
    events == "triple" ~ "triple", 
    events == "home_run" ~ "home_run", 
    TRUE ~ "other")) %>%
  filter(if_fielding_alignment %in% c("Infield shift", "Standard", "Strategic")) %>%
  filter(events_group %in% c("out", "single", "double", "triple", "home_run"), 
         launch_angle >= -20 & launch_angle <=50) %>%
  mutate(launch_angle_group = 
           cut(launch_angle, breaks = c(-20, -15, -10, -5, 0, 5, 10, 15, 20, 25, 
                                        30, 35, 40, 45, 50))) %>%
  filter(!is.na(launch_angle_group)) %>%
  ggplot(aes(x = launch_angle_group, fill = events_group))+
  geom_bar(position = "fill")+
  scale_fill_manual(values = pal2)+
  facet_wrap(~if_fielding_alignment, ncol = 1)+
  labs(x = "launch angle", y="", fill = "")+
  theme_minimal()
faceted_by_alignment
```
---
## Pitch type doesn't seem to have an impact either

```{r, fig.width = 14, fig.height = 6, echo = FALSE, fig.align='center'}
outcome_faceted <- batter_all_2019 %>%
  filter(description == "hit_into_play") %>%
  mutate(events_group = case_when(
    events %in% c("field_out", "other_out", "grounded_into_double_play", "double_play", 
                  "fielders_choice_out", "force_out", "sac_fly", "sac_fly_double_play", "sac_bunt_double_play", 
                  "field_error", "sac_fly", "fielders_choice", "triple_play") ~ "out",
    events == "single" ~ "single",
    events == "double" ~ "double", 
    events == "triple" ~ "triple", 
    events == "home_run" ~ "home run", 
    TRUE ~ "other")) %>%
  filter(events_group %in% c("out", "single", "double", "triple", "home run")) %>%
  mutate(pitch_group = case_when(
    pitch_name %in% c("4-Seam Fastball", "Fastball", "Cutter", "Sinker", "2-Seam Fastball") ~ "Fastball", 
    pitch_name %in% c("Changeup", "Split-Finger", "Screwball") ~ "Offspeed", 
    pitch_name %in% c("Slider", "Knuckle Curve", "Curveball", "Knuckle Curve", 
                      "Knuckleball", "Eephus") ~ "Breaking", 
    TRUE ~ pitch_name
  )) %>%
  filter(!is.na(pitch_group) & pitch_group %in% c("Offspeed", "Breaking", "Fastball")) %>%
  ggplot(aes(x=launch_angle, y = launch_speed, color = events_group)) +
  geom_point(alpha = 0.75) + 
  scale_x_continuous(n.breaks = 15, limits = c(-60, 90))+
  scale_color_manual(values = rev(pal2))+
  facet_wrap(~pitch_group, ncol = 1)+
  labs(x = "launch angle", y = "exit velocity (mph)", color = "")+
  theme_minimal()+
  theme(panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank())
outcome_faceted
```
---

### We can make wOBA-maximizing recommendations for changes in a player's launch angle based on their exit velocity 

.pull-left[
```{r, fig.width = 8, fig.height = 6, echo = FALSE, fig.align='center'}
wOBA_by_LA_EV <- batter_all_2019 %>%
  filter(description == "hit_into_play") %>%
  ggplot(aes(x=launch_angle, y = launch_speed, color = woba_value)) +
  geom_point(alpha = 0.5) + 
  scale_x_continuous(n.breaks = 15, limits = c(-60, 90))+
  scale_color_gradient(low = "darkblue", high = "darkorange")+
  labs(x = "launch angle", y = "exit velocity (mph)", 
       title = "wOBA Value by launch angle and exit velocity", color = "wOBA Value")+
  theme_minimal()+
  theme(
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank())

trout_stat_density <- batter_all_2019 %>%
  filter(player_name == "Trout, Mike", description == "hit_into_play") 

trout_density_2019 <- wOBA_by_LA_EV + stat_density2d(data = trout_stat_density, color = "white") + 
  labs(title = "Mike Trout")
trout_density_2019
```
]

.pull-right[
```{r, fig.width = 6, fig.height = 5.5, echo = FALSE, fig.align='center'}
heyward_stat_density <- batter_all_2019 %>%
  filter(player_name == "Heyward, Jason", description == "hit_into_play") 

heyward_density_2019 <- wOBA_by_LA_EV + stat_density2d(data = heyward_stat_density, color = "white") + 
  labs(title = "Jason Heyward")
heyward_density_2019
```
]
---
### Now how can we model these player recommendations? 

.pull-left[
- We will fit a **Generalized Additive Model, or GAM**, to **launch angle** and 
**exit velocity** with the target **wOBA**
  - GAMs have the ability to model complex *non-linear and non-monotonic* relationships between predictors and response variables, such as we see when we smooth the relationship between predictors **exit velocity** and **launch angle** with **wOBA**. 
]
.pull-right[
```{r, echo=FALSE, fig.width = 6, fig.height = 2.5}
batter_all_2019 %>%
  filter(description == "hit_into_play") %>%
  ggplot(aes(launch_angle, woba_value))+
  geom_smooth(se = FALSE)+
  scale_x_continuous(limits = c(-40, 60))+
  labs(title = "Launch angle vs. wOBA", x="", y="")+
  theme_minimal()+
  theme(plot.title.position = "plot", 
        plot.title = element_text(face = "bold"), 
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(), 
        panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_blank())
```

```{r, echo =FALSE, fig.width = 6, fig.height = 2.5}
batter_all_2019 %>%
  filter(description == "hit_into_play") %>%
  ggplot(aes(launch_speed, woba_value))+
  geom_smooth(se = FALSE)+
  scale_x_continuous(limits = c(50, 120))+
  labs(title = "Exit Velocity vs. wOBA", x="", y="")+
  theme_minimal()+
  theme(plot.title.position = "plot", 
        plot.title = element_text(face = "bold"), 
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(), 
        panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_blank())
```
]
---
### Our first model missed out on the relationship *between* launch angle and exit velocity in impacting wOBA

.pull-left[
**Actual relationship**
```{r, fig.height = 4, fig.width = 6, echo=FALSE}
batter_all_2019 %>%
  filter(description == "hit_into_play") %>%
  ggplot(aes(x=launch_angle, y=launch_speed, color = woba_value))+
  geom_point() + 
  labs(x="launch angle", y = "exit velocity", color = "wOBA", title = "wOBA by launch angle and exit velocity")+
  theme_minimal()+
  theme(plot.title.position = "plot", 
        plot.title = element_text(face = "bold"), 
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(), 
        panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_blank())
```

]
.pull-right[
**Our model's output**
```{r, echo=FALSE, fig.width=6, fig.height = 4}
woba_model <- gam(woba_value ~ s(launch_angle) + s(launch_speed), 
                  data = batter_all_2019, method = "REML")

made_up_data <- expand.grid(launch_angle = seq(-50,70,2), launch_speed = seq(50,110,2))
made_up_preds <- tibble(gam.preds = predict(woba_model, newdata = made_up_data)) 
test <- bind_cols(made_up_data, made_up_preds)

test %>%
  ggplot(aes(x=launch_angle, y = launch_speed, fill = gam.preds)) +
  geom_raster() + 
  labs(x="launch angle", y="exit velocity", 
       fill = "wOBA", 
       title = "Raster plot: gam(wOBA ~ s(launch angle) + s(exit velocity))")+
  theme_minimal()+
  theme(panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(), 
        panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_blank())
  
```
]
---
### With an interaction in our model, we much better see the true relationships, such as the increased wOBA from extra base hits and the band of singles

.pull-left[
**Actual relationship**
```{r, fig.height = 4, fig.width = 6, echo=FALSE}
batter_all_2019 %>%
  filter(description == "hit_into_play") %>%
  ggplot(aes(x=launch_angle, y=launch_speed, color = woba_value))+
  geom_point() + 
  labs(x="launch angle", y = "exit velocity", color = "wOBA", title = "wOBA by launch angle and exit velocity")+
  theme_minimal()+
  theme(plot.title.position = "plot", 
        plot.title = element_text(face = "bold"), 
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(), 
        panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_blank())
```
]
.pull-right[
**Our model's output** 
```{r, echo=FALSE, fig.width=6, fig.height = 4, message=FALSE, warning=FALSE}
woba_model2 <- gam(woba_value ~ s(launch_angle, launch_speed, k=200), data = batter_all_2019, 
                   method = "REML")

made_up_preds2 <- tibble(gam.preds = predict(woba_model2, newdata = made_up_data)) 
test2 <- bind_cols(made_up_data, made_up_preds2)

test2 %>%
  ggplot(aes(x=launch_angle, y = launch_speed, fill = gam.preds)) +
  geom_raster() + 
  labs(x="launch angle", y="exit velocity", 
       fill = "wOBA", title = "Raster plot: gam(wOBA ~ s(launch angle, exit velocity, k=200))")+
  theme_minimal()+
  theme(panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(), 
        panel.grid.major.x = element_blank(),
        panel.grid.major.y = element_blank())
  
```
]
---
### Next, we created a function that uses player data and our model to manually adjust an individual player's launch angle to best maximize their wOBA. 

##### Let's test this out for Jason Heyward. 

.pull-left[
**Our original thoughts...**
- As a reminder, based on **Jason Heyward's** density plot we recommended he try to increase his launch angle to around 10-15 degrees. 
  - This will increase Heyward's number of singles and doubles and decrease groundouts, increasing wOBA. 
```{r, fig.width = 6, fig.height = 5.5, echo = FALSE, fig.align='center'}
heyward_stat_density <- batter_all_2019 %>%
  filter(player_name == "Heyward, Jason", description == "hit_into_play") 

heyward_density_2019 <- wOBA_by_LA_EV + stat_density2d(data = heyward_stat_density, color = "white") + 
  labs(title = "Jason Heyward")
heyward_density_2019
```
]

.pull-right[

**What does our model tell us?**
```{r, echo = FALSE}
changing_launch_angle <- function(player_data, woba_model2, net_change) {
  #create tibble of predicted using model
  tibbletest <- tibble(gam.preds = predict(woba_model2, newdata = player_data))
  #find mean woba
  `predicted woba with changes` <- round(mean(tibbletest$gam.preds, na.rm = TRUE), 3)
  
  #begin by increasing launch angles by 1 to see what happens to mean 
  add_player_data <- player_data
  #launch_angle_original <- mean(player_data$launch_angle, na.rm = TRUE)
  add_player_data$launch_angle <- player_data$launch_angle + 1
  tibbletest <- tibble(gam.preds = predict(woba_model2, newdata = add_player_data))
  add_woba_mean <- mean(tibbletest$gam.preds, na.rm = TRUE)
  
  #begin by decreasing launch angles by 1 to see what happens to mean 
  subtract_player_data <- player_data
  subtract_player_data$launch_angle <- player_data$launch_angle - 1
  tibbletest <- tibble(gam.preds = predict(woba_model2, newdata = subtract_player_data))
  subtract_woba_mean <- mean(tibbletest$gam.preds, na.rm = TRUE)
  
  #if increasing the angles leads to a greater woba
  if (add_woba_mean > `predicted woba with changes`) {
    #increase the angles again
    net_change <- net_change + 1
    changing_launch_angle(add_player_data, woba_model2, net_change)
  } else if (subtract_woba_mean > `predicted woba with changes`) {
    #decrease the angles again 
    net_change <- net_change -1
    changing_launch_angle(subtract_player_data, woba_model2, net_change)
  } else {
    #no change 
    return (tibble(`predicted woba with changes`, `recommended launch angle` = round(mean(player_data$launch_angle, na.rm = TRUE),0), `degree change` = net_change))
  }
}

jason_heyward <- batter_all_2019 %>%
  filter(player_name == "Heyward, Jason", description == "hit_into_play")

jhey_recs <- changing_launch_angle(jason_heyward, woba_model2, 0)
knitr::kable(head(jhey_recs), format = 'html')
```

]

### What does our model recommend for Mike Trout?

.pull-left[
**Our original thoughts...**
- Based on **Mike Trout's** density plot, we didn't recommend any changes to his swing...he is 3x MVP after all! 
```{r, fig.width = 6, fig.height = 5.5, echo = FALSE, fig.align='center'}
trout_stat_density <- batter_all_2019 %>%
  filter(player_name == "Trout, Mike", description == "hit_into_play") 

trout_density_2019 <- wOBA_by_LA_EV + stat_density2d(data = trout_stat_density, color = "white") + 
  labs(title = "Mike Trout")
trout_density_2019
```
]
.pull-right[
```{r, echo=FALSE}
miketrout <- batter_all_2019 %>%
  filter(player_name == "Trout, Mike", description == "hit_into_play")

trout_recs <- changing_launch_angle(miketrout, woba_model2, 0)
knitr::kable(head(trout_recs), format = 'html')
```

]

